---
title: "Splitting and Resampling in R"
format:
  html:
    fontsize: 25px
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

The [Data Splitting](week2_3.qmd) page explained why we hold data back. This page does it in R.

::: {.callout-note icon="false"}
## Companion R Markdown File

All three pages of this unit are also available as one R Markdown file you can open in RStudio and run from start to finish. It builds its own data and includes exercises at the end.

[Download `modeling_pipeline.Rmd`](rmd/modeling_pipeline.Rmd)
:::

To follow along, load these three packages first. All of them came with the course setup, so you should not need to install anything.

```{r}
library(rsample)      # splitting and resampling
library(caret)        # fitting and cross validation
library(AmesHousing)  # the example data
```

::: {style="color: yellow"}
## The Ames Housing Data
:::

Everything on this page and the two that follow uses the **Ames housing** data, which the regression unit also uses.

-   Every sale of a residential property in Ames, Iowa between 2006 and 2010, collected by Dean De Cock for teaching.
-   **2,930 houses** and **81 variables**, so each row is one sale.
-   The target is `Sale_Price`, in dollars.
-   The predictors describe the house rather than the buyer: size, age, quality ratings, garage and basement details, and which of 28 neighborhoods it sits in.
-   About half the variables are categorical and half numeric, and there are no missing values, which is unusual and makes it a clean teaching dataset.

```{r}
ames <- make_ames()

dim(ames)
summary(ames$Sale_Price)
```

Prices run from about \$13,000 to \$755,000 with a median near \$160,000, and the mean sits above the median because a handful of expensive houses pull the distribution to the right. That skew matters later, both for stratified splitting and for the difference between RMSE and MAE.


::: {style="color: yellow"}
## Making the Split
:::

The `rsample` package handles splitting. Three functions do almost all of the work:

-   `initial_split()` decides which rows go where
-   `training()` pulls out the training rows
-   `testing()` pulls out the test rows

```{r}
set.seed(559)

split <- initial_split(ames, prop = 0.7)

ames_train <- training(split)
ames_test  <- testing(split)

c(train = nrow(ames_train), test = nrow(ames_test))
```

Two things to notice.

-   `prop = 0.7` puts 70 percent of the rows in training. There is nothing magic about 0.7. Common choices run from 0.6 to 0.8, and the tradeoff is the one from the previous page: more training data gives a better model, more test data gives a more stable estimate of how good it is.
-   `set.seed()` comes first. The split is random, so without a seed you get a different split every run and none of your numbers reproduce.

::: {style="color: lightblue"}
### Stratified Splitting
:::

A plain random split can get unlucky. If expensive houses land mostly in one half, the two halves are not comparable.

`strata` fixes this by splitting *within* groups of the outcome and then combining.

```{r}
set.seed(559)

split <- initial_split(ames, prop = 0.7, strata = "Sale_Price")

ames_train <- training(split)
ames_test  <- testing(split)

c(train_mean = mean(ames_train$Sale_Price),
  test_mean  = mean(ames_test$Sale_Price))
```

The two means now sit close together by construction rather than by luck.

Stratifying matters most when:

-   The outcome is skewed, as sale prices are.
-   The outcome is a category with a rare level. If only 8 percent of patients relapse, an unstratified split can easily leave you with very few relapses in the test set.

::: {.callout-tip icon="false"}
## Try It

Run the unstratified split a few times with different seeds and watch the training and test means move around. Then do the same with `strata = "Sale_Price"`. How much does stratifying reduce the wobble?
:::

::: {style="color: yellow"}
## Cross Validation in Code
:::

The previous page described k fold cross validation in six steps. In `caret` those six steps are one function call.

`trainControl()` describes *how* to resample. `train()` does the fitting.

```{r}
cv <- trainControl(
  method = "cv",
  number = 10
)
```

That says: split the training data into 10 folds, hold each fold out in turn, fit on the other 9, and score on the held out fold.

Now fit a model using it.

```{r}
set.seed(559)

fit_cv <- train(
  Sale_Price ~ Gr_Liv_Area,
  data      = ames_train,
  method    = "lm",
  trControl = cv
)

fit_cv
```

The reported RMSE is the **average across the 10 folds**, not a single number from a single split. That is the whole point: one holdout split gives an unstable estimate, and averaging ten of them gives a steadier one.

::: {style="color: lightblue"}
### What Just Happened
:::

It is worth being explicit, because `train()` hides a lot:

1.  The training data were divided into 10 folds.
2.  The model was fit 10 separate times, each time on 9 folds.
3.  Each fit was scored on the fold it had not seen.
4.  The 10 scores were averaged.
5.  Finally, a model was fit once on all of the training data. That is the model you keep.

Note what did **not** happen. The test set was never touched. It is still sitting untouched in `ames_test`, which is exactly where it should stay until the very end.

::: {.callout-tip icon="false"}
## Quick Check

1.  Why does `set.seed()` go before `initial_split()` rather than after it?
2.  Counting everything in the list above, how many times was the model actually fit?
:::


::: {style="color: lightblue"}
### Other Resampling Methods
:::

`method` accepts more than `"cv"`:

| Setting | What it does |
|----|----|
| `"cv"` | k fold cross validation |
| `"repeatedcv"` | k fold, repeated several times with different fold assignments |
| `"boot"` | bootstrap resampling, sampling with replacement |
| `"LOOCV"` | leave one out, where k equals the number of rows |
| `"none"` | no resampling, just fit once |

`"repeatedcv"` is useful when your dataset is small enough that a single set of folds is itself noisy:

```{r}
cv_rep <- trainControl(method = "repeatedcv", number = 10, repeats = 5)
```

::: {.callout-note icon="false"}
## A Note on Two Toolkits

You will see two different styles in this course and in the wild.

-   **caret** is the older framework. One function, `train()`, does the fitting, and `trainControl()` configures it.
-   **tidymodels** is the newer family, which splits the same job across several packages: `rsample` for splitting, `recipes` for preprocessing, `parsnip` for models, and `yardstick` for metrics.

We use `rsample` for splitting and `caret` for fitting, which is a common and perfectly reasonable mix. They are not rivals you must choose between; they are two vocabularies for the same pipeline.
:::

Next: [Evaluating Model Performance](p02_evaluating_performance.qmd), which explains what that RMSE number actually measures.
