---
title: "A Complete Pipeline"
format:
  html:
    fontsize: 25px
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

We now have all the pieces. This page runs them end to end, once, on the Ames housing data.

To follow along, load these five packages.

```{r}
library(rsample)      # splitting
library(recipes)      # preprocessing
library(caret)        # fitting and cross validation
library(AmesHousing)  # the example data
library(dplyr)        # the pipe used in the recipe below
```

The pipeline is six steps:

1.  Split the data
2.  Write a preprocessing recipe
3.  Choose a resampling scheme
4.  Fit and tune
5.  Compare candidate models
6.  Evaluate on the test set, exactly once

::: {style="color: yellow"}
## Step 1: Split
:::

```{r}
ames <- make_ames()

set.seed(559)
split <- initial_split(ames, prop = 0.7, strata = "Sale_Price")

ames_train <- training(split)
ames_test  <- testing(split)
```

From this point until step 6, `ames_test` does not appear anywhere.

::: {style="color: yellow"}
## Step 2: A Preprocessing Recipe
:::

Most data need work before modelling. A **recipe** is a written list of those steps that is *learned from the training data and then applied elsewhere*.

```{r}
rec <- recipe(Sale_Price ~ Gr_Liv_Area + Year_Built + Neighborhood,
              data = ames_train) %>%
  step_other(all_nominal_predictors(), threshold = 0.05) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())

rec
```

Reading it in order:

-   `step_other` lumps rare neighborhoods into a single "other" category, since a neighborhood appearing four times cannot support its own estimate.
-   `step_normalize` centers and scales the numeric predictors.
-   `step_dummy` converts categories into 0 and 1 indicator columns.
-   `step_zv` drops any predictor with zero variance.

::: {.callout-important icon="false"}
## Why a Recipe Rather Than Just Doing It

Suppose you center and scale `Gr_Liv_Area` before splitting. The mean you subtract is computed from **all** the data, including the rows destined for the test set. Information from the test set has now leaked into training, and your test estimate is optimistic.

A recipe prevents this by construction. It records that a mean must be subtracted, learns that mean from the training data only, and applies that same stored value to any new data later. The rule travels with the model.

This is data leakage, which the [Data Splitting](week2_3.qmd) page listed as a hazard. Preprocessing before splitting is by far the most common way it happens, and it is invisible: nothing errors, your numbers just quietly become too good.
:::

::: {style="color: yellow"}
## Step 3: Resampling Scheme
:::

```{r}
cv <- trainControl(method = "cv", number = 10)
```

::: {style="color: yellow"}
## Step 4: Fit
:::

Passing the recipe to `train()` rather than a formula means the recipe is re-learned inside every fold, on that fold's training portion only. That is what keeps the leakage protection honest during cross validation.

```{r}
set.seed(559)

fit_lm <- train(
  rec,
  data      = ames_train,
  method    = "lm",
  trControl = cv,
  metric    = "RMSE"
)

fit_lm$results
```

::: {style="color: yellow"}
## Step 5: Compare Candidates
:::

The no free lunch theorem from [Important Terms and Distinctions](week2_1.qmd) said we cannot know in advance which method wins. So we try more than one and compare them on the same folds.

Here is a k nearest neighbors model with a hyperparameter to tune. `tuneGrid` lists the values to try, and cross validation picks among them.

```{r}
set.seed(559)

fit_knn <- train(
  rec,
  data      = ames_train,
  method    = "knn",
  trControl = cv,
  metric    = "RMSE",
  tuneGrid  = data.frame(k = c(5, 10, 15, 20, 25))
)

fit_knn$results[, c("k", "RMSE", "Rsquared", "MAE")]
```

Each row is one value of `k`, scored by cross validation. `train()` keeps the best:

```{r}
fit_knn$bestTune
```

Now compare the two models. Because both used the same seed and the same folds, this is a fair comparison.

```{r}
results <- resamples(list(linear = fit_lm, knn = fit_knn))
summary(results)$statistics$RMSE
```

::: {.callout-tip icon="false"}
## Quick Check

1.  Why is the recipe handed to `train()` rather than applied to `ames_train` beforehand?
2.  Both models were compared on the same folds. Why does using the *same* folds matter?
:::


::: {.callout-tip icon="false"}
## Notice What Chose the Winner

The validation folds picked the value of `k`, and the validation folds picked between the two models. Both are decisions made *with data*, which is precisely why they cannot be made using the test set. Every choice you make by looking at a number is a way of fitting to that data.
:::

::: {style="color: yellow"}
## Step 6: The Test Set, Once
:::

Pick the winner, then look at the test data for the first and only time.

```{r}
final_model <- fit_lm

test_predictions <- predict(final_model, ames_test)

postResample(pred = test_predictions, obs = ames_test$Sale_Price)
```

That is the number you report. It is an estimate of how the model will perform on houses nobody has seen.

::: {.callout-important icon="false"}
## The Rule

If you now go back, change the model, and re-run this, the number stops meaning what it meant. The test set has become part of model selection, and your estimate is optimistic again.

You get one look. This is why steps 1 through 5 exist.
:::

::: {style="color: yellow"}
## The Whole Thing
:::

Every pipeline in this course has this shape, whatever the model:

```{r, eval=FALSE}
# 1. split
split <- initial_split(data, prop = 0.7, strata = "outcome")
train <- training(split)
test  <- testing(split)

# 2. recipe
rec <- recipe(outcome ~ ., data = train) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors())

# 3. resampling
cv <- trainControl(method = "cv", number = 10)

# 4 and 5. fit, tune, compare
fit <- train(rec, data = train, method = "lm",
             trControl = cv, metric = "RMSE")

# 6. once
postResample(predict(fit, test), test$outcome)
```

From here the course mostly changes one thing at a time. The regression unit changes what goes in `method`. The feature engineering unit changes what goes in the recipe. The tree unit changes both. The pipeline stays the same.
