---
title: "Working with Data Frames"
format:
  html:
    fontsize: 25px
editor_options:
  chunk_output_type: console
---

::: {style="color: yellow"}
## Data Frames
:::

Nearly everything we do this semester will start with a data frame: a rectangular table where each row is a case (a person, a customer, a house) and each column is a variable. In this section we'll practice pulling pieces out of a data frame, creating new variables, and chaining operations together. These are the skills you'll use in every single lab and assignment.

Let's read in the marketing data we saw previously (download: [ifood_df.csv](data/ifood_df.csv)).

```{r}
df <- read.csv("data/ifood_df.csv")
```

A few functions are useful for getting oriented whenever you load a new dataset:

```{r}
dim(df)      # number of rows and columns
nrow(df)     # number of rows (cases)
names(df)    # variable names
```

```{r}
head(df[, 1:6])  # first few rows of the first 6 columns
```

::: {style="color: yellow"}
## Selecting a Variable with `$`
:::

The `$` operator pulls a single variable out of a data frame as a vector. This is one of the most common operations in R, and you will see it constantly in this course.

```{r}
mean(df$Income)    # average household income
table(df$Complain) # how many customers lodged a complaint?
```

We can also use `$` to create a new variable in the data frame. For example, the number of children at home is split across two variables, `Kidhome` and `Teenhome`. Let's combine them:

```{r}
df$TotalKids <- df$Kidhome + df$Teenhome
table(df$TotalKids)
```

::: {style="color: yellow"}
## Subsetting with `[rows, columns]`
:::

Square brackets let us take a subset of a data frame using the pattern `df[rows, columns]`. Leaving either slot empty means "keep everything."

```{r}
df[1:5, ]                        # first five rows, all columns
head(df[, c("Income", "Age")])   # all rows, two named columns
```

We can also select rows using a logical condition. For example, customers with incomes above \$75,000:

```{r}
high_income <- df[df$Income > 75000, ]
nrow(high_income)
```

Read that as: "keep the rows of `df` where `df$Income > 75000` is `TRUE`, and all columns."

::: {style="color: yellow"}
## Missing Values: `NA`
:::

Real data is rarely complete. R represents a missing value with `NA` ("not available"), and `NA` behaves differently than you might expect: most computations involving a missing value return a missing value.

```{r}
x <- c(4, 7, NA, 2)
mean(x)
```

That `NA` answer is R being honest: it can't know the mean of numbers it doesn't have. If you've decided that ignoring the missing values is appropriate, say so explicitly with `na.rm = TRUE`:

```{r}
mean(x, na.rm = TRUE)
```

To find missing values, use `is.na()`. Asking `x == NA` does *not* work, because comparing anything to an unknown is also unknown:

```{r}
is.na(x)        # which entries are missing?
sum(is.na(x))   # how many are missing?
```

For a whole data frame, `colSums(is.na(df))` shows the missing count per variable. Our marketing data happens to be a cleaned dataset:

```{r}
sum(is.na(df))  # no missing values here
```

Real psychological data will rarely be this kind; the General Social Survey data in your first assignment certainly won't be. Handling missingness *well* is a modeling topic we'll treat properly in the feature engineering unit; for now, the essential skills are noticing missing values and understanding why `NA`s propagate through your calculations.

::: {style="color: yellow"}
## The Pipe Operator `%>%`
:::

As our analyses grow, we often want to perform several operations in sequence. The **pipe** operator `%>%` takes the object on its left and passes it to the function on its right. These two lines do exactly the same thing:

```{r, message=FALSE}
library(dplyr)

head(df[, 1:4])      # without the pipe
df[, 1:4] %>% head() # with the pipe
```

That looks like a small change, but pipes let us read a chain of steps top-to-bottom like a recipe, instead of inside-out. You will see `%>%` throughout the course materials, especially once we start building preprocessing and modeling pipelines.

*Note:* newer versions of R also have a built-in pipe written `|>`. It works the same way for our purposes, and you may see either form in the wild.

::: {style="color: yellow"}
## Data Wrangling with `dplyr`
:::

The `dplyr` package (loaded above) provides a set of verbs for working with data frames that pair naturally with the pipe. The four you'll see most often:

-   `select()`: choose columns

-   `filter()`: choose rows matching a condition

-   `mutate()`: create new variables

-   `summarise()`: collapse rows into summary values (often with `group_by()`)

```{r}
df %>% select(Income, Age, MntWines) %>% head()
```

```{r}
df %>% filter(Income > 75000) %>% nrow()
```

```{r}
df %>%
  mutate(TotalKids = Kidhome + Teenhome) %>%
  select(Income, TotalKids) %>%
  head()
```

Chaining verbs together is where this style shines. For example: among customers who did and did not complain, what is the average wine spending and group size?

```{r}
df %>%
  group_by(Complain) %>%
  summarise(
    avg_wine = mean(MntWines),
    n = n()
  )
```

::: {style="color: yellow"}
## Two Dialects, One Language
:::

Base R subsetting (`$`, `[ , ]`) and `dplyr` verbs are two ways of doing the same things, and this course uses both: base R for quick one-liners, `dplyr` chains for multi-step preprocessing. You should be comfortable *reading* both. When writing your own code, use whichever feels clearer to you.
