---
title: "Feature Filtering"
format: 
  html: 
    fontsize: 25px
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

::: {style="color: yellow"}
## Setup
:::

To follow along, load these packages and the dataset we assembled on the [Example Data](week2_5b.qmd) page. If you did not build it yourself, you can [download `data.RDS`](data/data.RDS) and save it in a `data` folder next to your script.

```{r}
library(dplyr)  # data manipulation and the pipe
library(caret)  # nearZeroVar(), used below to flag low variance features

data <- readRDS("data/data.RDS")
```

Feature filtering is a feature selection technique in machine learning where features are evaluated prior to the model fitting based on statistical or heuristic criteria. Features are then kept or discarded based on those criteria.

It is called filtering because it is a preprocessing step, part of a feature engineering pipeline, done before we actually train the model.

Typically, the goals of feature filtering are:

1.  **Reduce dimensionality:** Fewer features means simpler models, faster training, and less risk of overfitting.

2.  **Remove irrelevant or redundant features:** Avoids feeding the model noisy or useless inputs.

3.  **Improve interpretability:** Models are easier to understand when they focus only on meaningful features.

4.  **Boost performance:** Better generalization on new data when unnecessary features are excluded.

In practice, one of the filtering tasks we will typically conduct involves weeding out low variance features.

::: {style="color:orange"}
## Removing Low-Variance Features {style="color:orange"}
:::

Zero and near-zero variance variables are low-hanging fruit to eliminate.

-   **Zero Variance Features**: variable only contains a single unique value

-   **Near-Zero Variance Features**: variable contains only a few unique values

Zero and near-zero variance variables typically offer little to no information for model building. Furthermore, resampling (data-splitting) further complicates this picture because a given fold or sample may only contain a single value if the variable itself only contains a few unique values.

@handson suggest the following rule-of-thumb for removing low-variance features:

**Remove a variable if:**

-   The fraction of unique values over the sample size is low (e.g. 10%).

-   The ratio of the frequency of the most prevalent value to the frequency of the second most prevalent value is large (e.g. ≥20).

We can use the `caret` [@caret] package in R to look at these different metrics for our example data.

```{r}
# nearZeroVar() checks every column against the two rules above.
#   saveMetrics = TRUE returns the diagnostics for every variable. Without it
#   you get back only the column numbers of the ones it would drop, which is
#   useful for code but useless for understanding what is going on.
caret::nearZeroVar(data, saveMetrics = TRUE) %>%

  # the variable names arrive as row names rather than as a real column,
  # which makes them awkward to read, so this moves them into a column
  tibble::rownames_to_column()
```

Four columns come back. The two that matter are the ones from the rule of thumb above.

-   `freqRatio` is the frequency of the most common value divided by the frequency of the second most common. A big number means one value dominates the variable.
-   `percentUnique` is how many distinct values the variable has, as a percentage of the number of rows. A small number means the variable is close to constant.
-   `zeroVar` and `nzv` are `TRUE` or `FALSE` flags. They are `caret` applying the thresholds for you and saying whether it would drop the variable.

In a real pipeline you would not run this by hand. You would add `step_nzv()` to a recipe and let it happen inside the resampling, for the reasons the [Overview](week2_5a.qmd) gave. Running it here is just so you can see what the decision is based on.
