---
title: "Numeric Feature Engineering"
format: 
  html: 
    fontsize: 25px
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

::: {style="color: yellow"}
## Setup
:::

To follow along, load these packages and the dataset we assembled on the [Example Data](week2_5b.qmd) page. If you did not build it yourself, you can [download `data.RDS`](data/data.RDS) and save it in a `data` folder next to your script.

```{r}
library(dplyr)       # data manipulation and the pipe
library(tidymodels)  # the recipes package, used for every step below

data <- readRDS("data/data.RDS")
```

::: {style="color: yellow"}
## Numeric Feature Engineering
:::

Feature engineering on continuous or numeric features involves transforming the raw data to make it more useful for modeling. Common motivations include capturing nonlinear relationships (e.g., log or polynomial transforms), improving distributional properties (reducing skew or outlier impact), and putting features on comparable scales through standardization.

In some cases, continuous variables are binned into categories or combined into interactions to highlight patterns. Overall, these transformations help models detect structure in the data more effectively and improve both performance and interpretability.

::: {style="color: orange"}
### Skewness
:::

Parametric models with distributional assumptions (e.g., GLMs, and some regularized models) can benefit from minimizing the skewness of numeric features. One popular transformation is the [Yeo-Johnson transformation](https://en.wikipedia.org/wiki/Power_transform#Yeo%E2%80%93Johnson_transformation).

The Yeo-Johnsontransformation is a statistical technique used to stabilize variance and make data more normally distributed, similar to the Box-Cox transformation but more flexible. Unlike Box-Cox, it can handle both positive and negative values, making it useful for real-world data that span zero.

```{r}
# all_numeric_predictors() applies the transformation to the numeric features
# and leaves the outcome alone. A recipe only records the plan at this stage,
# so nothing is actually transformed until the recipe is prepped.
recipe(happy ~ ., data = data) %>%
  step_YeoJohnson(all_numeric_predictors())
```

::: {style="color: orange"}
### Standardization
:::

We will often want to standardize variables prior to our model fitting to put them on a common scale, typically with mean of zero and a standard deviation of one.

This prevents features with larger numeric ranges (e.g., income in dollars vs. age in years) from dominating distance-based methods (like k-NN, SVMs, or clustering). It can also help with optimization problems, may improve convergence and can help ensure that regularization penalties (like in ridge or lasso regression) are applied fairly across predictors.

```{r}
# step_center subtracts the mean, step_scale divides by the standard
# deviation. Run together they standardize each feature. Both learn those
# numbers from the training data alone, which is what keeps information
# from the test set out of the calculation.
recipe(happy ~ ., data = data) %>%
  step_center(all_numeric_predictors()) %>%
  step_scale(all_numeric_predictors())
```
