---
title: "R Markdown and Course Setup"
format:
  html:
    fontsize: 25px
editor_options:
  chunk_output_type: console
---

::: {style="color: yellow"}
## R Markdown
:::

In-class exercises and assignments in this course are distributed as **R Markdown** (`.Rmd`) files. An R Markdown file mixes written text with runnable R code. You will complete the code, **knit** the document into a polished HTML or PDF report, and submit it on Canvas. Learning this workflow now means the first graded exercise will hold no surprises.

### Creating an R Markdown Document

In RStudio, choose **File \> New File \> R Markdown...**, give it a title, and click OK. RStudio opens a template document you can knit immediately — try it before changing anything.

### The Anatomy of an R Markdown File

An `.Rmd` file has three ingredients:

1.  **The YAML header** — the block between `---` lines at the very top. It stores the title, author, date, and output format.

2.  **Prose** — ordinary text, with light formatting: `**bold**`, `*italics*`, `# headings`.

3.  **Code chunks** — R code between ```` ```{r} ```` and ```` ``` ```` fences. Insert a new chunk with the toolbar button or the shortcut **Cmd+Option+I** (Mac) / **Ctrl+Alt+I** (Windows).

You can run a single chunk with the green "play" button on its right edge — useful for checking your work as you go.

### Knitting

The **Knit** button at the top of the editor runs the entire document *top to bottom in a fresh R session* and produces the output file. That "fresh session" detail explains the most common student pitfalls:

-   **"It works in my console but won't knit."** Objects you created in the console don't exist when knitting. Every object your document uses must be created *by code chunks in the document*, in order.

-   **Never put `install.packages()` in a code chunk.** Install packages once, from the console. A document should *load* packages with `library()`, not install them.

-   **Keep the `.Rmd` and any data files in the same folder**, and refer to data by filename (`read.csv("mydata.csv")`). Knitting looks for files relative to where the `.Rmd` lives.

::: {style="color: yellow"}
## Install the Course Packages
:::

Throughout the semester we will lean on a set of R packages for wrangling, visualization, and machine learning. Install them all now by pasting this into your **console** (not an R Markdown chunk!):

```{r, eval=FALSE}
install.packages(c(
  "tidyverse",    # data wrangling and ggplot2 graphics
  "psych",        # describing data
  "tidymodels",   # modeling framework (includes rsample, recipes)
  "caret",        # classic ML toolkit
  "rpart",        # decision trees
  "ipred",        # bagging
  "ranger",       # random forests
  "gbm",          # gradient boosting
  "glmnet",       # regularized regression
  "vip",          # variable importance plots
  "AmesHousing",  # housing data used in the regression unit
  "cluster",      # clustering methods
  "factoextra"    # clustering and PCA visualizations
))
```

This can take a while — `tidyverse`, `tidymodels`, and `caret` are large. Start it, get a coffee, and let it finish. Do this **before the first lab** so class time isn't spent watching progress bars.

A few assignment-specific packages (for example, `gssr` for working with the General Social Survey) will be installed later with instructions provided in the assignment itself.

### Check Your Setup

If this runs without errors, you are ready for the semester:

```{r, eval=FALSE}
library(tidyverse)
library(tidymodels)
library(caret)
library(psych)
```

If something fails to load, re-run `install.packages()` for that package and read the error message — and if you're stuck, bring the error to class or office hours. Getting your machine working is a course goal in week one, not an obstacle.
