---
title: "Example Data"
format: 
  html: 
    fontsize: 25px
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

::: {style="color: yellow"}
## Example Data
:::

To demonstrate feature engineering we will use some example data from the [The General Social Survey (GSS)](https://gss.norc.org/).

GSS is a long-running, nationally representative survey of adults in the United States that has been conducted almost every two years since 1972 by the [National Opinion Research Center (NORC)](https://www.norc.org/) at the University of Chicago. The GSS data is often used to measure American's attitudes, behaviors, and beliefs on a wide range of topics, such as politics, religion, crime, race relations, family, work, and technology.

::: {.callout-note icon="false"}
## Further Information

A video introduction to the General Social Survey, covering what it is and how it is run.

[Introduction to the GSS](https://www.youtube.com/watch?v=jBb2mCmCcOk)
:::

There are two ways to access the GSS data. One is using the [GSS Data Explorer](#0). Another option is using the `gssr` [@gssr] R package. Here we will use the `gssr` [@gssr] package to download some example data.

To follow along, load these two packages first.

```{r}
library(gssr)   # downloading the GSS data
library(dplyr)  # data manipulation and the pipe
```

We will start by choosing `happy` as our target variable. This comes from Question 157 of the 2024 GSS where respondents were asked:

> Taken all together, how would you say things are these days - would you say that you are very happy, pretty happy, or not too happy?

Responses were coded such that 1 indicates "very happy", 2 indicates "pretty happy", and 3 indicates "not too happy", while NA indicates "don’t know."

We can also identify a bunch of features we think predict self-reported happiness and save our final dataset.

```{r}
# Downloads the full 2024 GSS. This is a big file, so it takes a moment.
# The result is a data frame with one row per respondent.
gss24 <- gss_get_yr(2024)

# The variable we want to predict. Storing it in an object rather than
# typing "happy" everywhere means we only have to change it in one place.
target <- "happy"

features <- c(
  "age",     # age of participant
  "sex",     # sex of participant
  "race",    # race of participant
  "educ",    # highest education completed by participant
  "income",  # income of participant
  "childs",  # number of children participant has
  "wrkstat", # work force status
  "marital", # marital status
  "born",    # whether or not participant was born in USA
  "partyid", # political party of participant
  "adults",  # num of fam members 18 or older
  "earnrs"   # number of earners in family  
)

# Square brackets subset a data frame as [rows, columns]. Leaving the rows
# slot empty means "keep every row". The columns slot keeps only our target
# and our features, so we go from around a thousand columns down to 13.
data <- gss24[, c(target, features)]

# table() counts how many times each value appears. useNA = "always" forces
# it to show the count of missing values too, which it would otherwise hide.
# Look at that NA column, because missing data is the subject of the next page.
table(data$happy, useNA = "always")
```

Next, let's clean up `data` a bit, discarding some of the labels and missing value information we don’t need. The data in `gss24` retains the labeling structure provided by the GSS. Variables are stored numerically with labels attached to them. Often, when using the data in R, it will be convenient to convert the categorical variables we are interested in to character or factor type instead.

Here we can use code from the [gssr package introduction](https://kjhealy.github.io/gssr/articles/overview.html) to simplify this recoding. The only thing we need to do is define the categorical variables in our data.

```{r}
# Two lists of column names. c() just glues values together into a vector.
# We write these out once so the code below can refer to a whole group of
# columns by name instead of repeating the same operation seven times.
cat_vars <- c("sex","race", "educ", "wrkstat", "marital", "born", "partyid")
cont_vars <- c("happy", "age", "income", "childs", "adults", "earnrs")

# Tidies up label text. It splits a label on spaces, uppercases the first
# letter of each word, lowercases the rest when strict = TRUE, then pastes the
# words back together. This is what turns "NEVER MARRIED" into "Never Married".
capwords <- function(x, strict = FALSE) {
    cap <- function(x) paste(toupper(substring(x, 1, 1)),
                  {x <- substring(x, 2); if(strict) tolower(x) else x},
                             sep = "", collapse = " " )
    sapply(strsplit(x, split = " "), cap, USE.NAMES = !is.null(names(x)))
}

# mutate() applies each of the four steps below to the columns named in it.
# across() is what lets one line act on a whole set of columns at once.
data <- data |>
  mutate(

    # The GSS does not leave unanswered questions blank. It fills them with
    # special codes for things like "Don't know" and "No answer". zap_missing
    # turns every one of those codes into a plain NA so the rest of R treats
    # them consistently. This has to run first. If we converted the columns
    # below before clearing these codes, the codes would survive as if they
    # were real answers.
    across(everything(), haven::zap_missing),

    # The continuous variables arrive carrying the GSS labelling structure.
    # as.numeric strips that off and leaves plain numbers we can do arithmetic
    # on, such as taking a mean.
    across(all_of(cont_vars), as.numeric),

    # The categorical variables get the opposite treatment. as_factor converts
    # the stored numeric codes into a factor using the labels the GSS ships
    # with the data, so a stored 1 becomes its written category instead.
    across(all_of(cat_vars), forcats::as_factor),

    # Those labels arrive in all capitals. fct_relabel runs every level of the
    # factor through capwords to clean them up. The \(x) is shorthand for
    # function(x), which we need here only because we want to pass capwords its
    # strict = TRUE argument.
    across(all_of(cat_vars), \(x) forcats::fct_relabel(x, capwords, strict = TRUE))
  )
```

```{r,echo=F,results='hide'}
saveRDS(data,"data/data.RDS")
```
