---
title: "Multiple Linear Regression"
format:
  html:
    fontsize: 25px
editor_options:
  chunk_output_type: console
bibliography: references.bib
---

[Simple Linear Regression](week3_1.qmd) predicted second semester grade from a single predictor. This page adds more predictors. Adding predictors never makes the fit to the training data worse, so the question is whether the model **predicts** better.

::: {style="color: yellow"}
## Setup
:::

```{r}
library(dplyr)    # data manipulation and the pipe
library(ggplot2)  # plots
library(caret)    # train() for fitting and cross validation
library(rsample)  # splitting
library(vip)      # variable importance plots
```

We continue with the student dropout data from the previous page. If you did not build it there, [download `dropout.csv`](data/dropout.csv).

```{r}
dropout <- read.csv("data/dropout.csv", sep = ";") %>%
  rename(
    grade_sem1      = Curricular.units.1st.sem..grade.,
    grade_sem2      = Curricular.units.2nd.sem..grade.,
    units_sem1      = Curricular.units.1st.sem..approved.,
    admission_grade = Admission.grade,
    prev_grade      = Previous.qualification..grade.,
    age             = Age.at.enrollment,
    scholarship     = Scholarship.holder,
    fees_current    = Tuition.fees.up.to.date
  )

# Same decision as the previous page: keep students who completed units,
# so the model is not just separating completers from non-completers
completers <- dropout %>% filter(grade_sem1 > 0, grade_sem2 > 0)

nrow(completers)
```

::: {style="color: yellow"}
## Overview of Multiple Linear Regression
:::

Multiple linear regression extends the simple linear regression framework to include more than one predictor:

-   **more than one feature** (also called the predictors or independent variables),
-   **one target** (also called the outcome or dependent variable).

It models the relationship between a single outcome $Y_i$ and multiple predictors $X_{1i}, X_{2i}, \dots, X_{pi}$ for each observation $i = 1, \dots, n$.

::: {style="color: yellow"}
## The Model
:::

$$
Y_i = \beta_0 + \beta_1 X_{1i} + \beta_2 X_{2i} + \dots + \beta_p X_{pi} + \varepsilon_i
$$

where:

-   $Y_i$: the outcome for observation $i$.
-   $X_{ji}$: the value of predictor $j$ for observation $i$.
-   $\beta_0$: the intercept, the expected value of $Y_i$ when all $X_{ji} = 0$.
-   $\beta_j$: the coefficient for predictor $X_{ji}$, the expected change in $Y_i$ for a one unit increase in $X_{ji}$, holding all other predictors constant.
-   $\varepsilon_i$: the error term, capturing variation not explained by the predictors.

::: {style="color: yellow"}
## Estimating the Model
:::

The coefficients are estimated by ordinary least squares, which chooses the values minimizing the residual sum of squares:

$$
\text{RSS} = \sum_{i=1}^n (y_i - \hat{y}_i)^2
$$

This is the same loss function idea from the previous page. Nothing about the learning procedure changes when you add predictors. There are simply more coefficients to search over.

::: {style="color: yellow"}
## Implementation in R
:::

Predictors are joined with `+` on the right hand side of the formula.

```{r}
fit_mlr <- lm(grade_sem2 ~ grade_sem1 + admission_grade, data = completers)

coef(fit_mlr)
```

::: {style="color: lightblue"}
### Interpreting Coefficients
:::

Each $\beta_j$ reflects the **unique contribution** of its predictor, after adjusting for the others.

```{r}
b <- coef(fit_mlr)
```

-   The intercept, `r round(b[1], 2)`, is the predicted second semester grade for a student with a first semester grade of 0 and an admission grade of 0. Neither is possible, so treat it as the number the equation needs rather than a claim about any student.
-   The coefficient on `grade_sem1`, `r round(b[2], 3)`, is the predicted change in second semester grade for a one point increase in first semester grade, **holding admission grade constant**.
-   The coefficient on `admission_grade`, `r round(b[3], 3)`, is the predicted change for a one point increase in admission grade, holding first semester grade constant.

Admission grade on its own is related to second semester grade, but much of what it tells you is already contained in the first semester grade. Its coefficient reports only what it adds.

::: {.callout-warning icon="false"}
## Holding Constant Is Not an Intervention

"Holding admission grade constant" describes a comparison between students who already differ. It does not describe what would happen if you changed somebody's admission grade.

A coefficient is a statement about an association in this dataset. Turning it into a claim about what an intervention would do requires assumptions this model cannot supply.
:::

::: {style="color: lightblue"}
### Interactions
:::

An interaction occurs when the effect of one predictor depends on the level of another. In other words, one variable is moderated by another.

$$
Y_i = \beta_0 + \beta_1 X_{1i} + \beta_2 X_{2i} + \beta_3 (X_{1i} \cdot X_{2i}) + \varepsilon_i
$$

where $\beta_3$ captures the interaction, and $(X_{1i} \cdot X_{2i})$ is the product of the two predictors. If $\beta_3 \neq 0$, the effect of $X_1$ on $Y$ changes depending on $X_2$.

Without an interaction, the model assumes each predictor contributes independently, which may oversimplify reality.

```{r}
# The colon creates the product term
fit_interaction <- lm(
  grade_sem2 ~ grade_sem1 + admission_grade + grade_sem1:admission_grade,
  data = completers
)

coef(fit_interaction)
```

::: {style="color: yellow"}
## Assessing Model Accuracy
:::

In this course the goal is prediction, so we return to the splitting and cross validation framework from [the pipeline unit](p01_splitting_resampling.qmd).

```{r}
set.seed(123)

split <- initial_split(completers, prop = 0.7, strata = "grade_sem2")
train  <- training(split)
test   <- testing(split)

c(train = nrow(train), test = nrow(test))
```

```{r}
# 10 fold cross validation, used identically for every model below so the
# comparison between them is fair
ctrl <- trainControl(method = "cv", number = 10)
```

We compare four models of increasing complexity.

::: {style="color: lightblue"}
### Model 1: One Predictor
:::

```{r}
set.seed(123)
cv_model1 <- train(grade_sem2 ~ grade_sem1,
                   data = train, method = "lm", trControl = ctrl)

cv_model1$results
```

```{r}
#| echo: false
r1 <- min(cv_model1$results$RMSE)
```

The cross validated RMSE is `r round(r1, 3)`. Read it as: on students it has not seen, this model's predictions are off by about `r round(r1, 2)` grade points on average.

::: {style="color: lightblue"}
### Model 2: Adding Admission Grade
:::

```{r}
set.seed(123)
cv_model2 <- train(grade_sem2 ~ grade_sem1 + admission_grade,
                   data = train, method = "lm", trControl = ctrl)

cv_model2$results
```

```{r}
#| echo: false
r2 <- min(cv_model2$results$RMSE)
```

RMSE is now `r round(r2, 3)`, an improvement of about `r round(r1 - r2, 3)` grade points.

::: {style="color: lightblue"}
### Model 3: Adding the Interaction
:::

```{r}
set.seed(123)
cv_model3 <- train(grade_sem2 ~ grade_sem1 + admission_grade + grade_sem1:admission_grade,
                   data = train, method = "lm", trControl = ctrl)
```

```{r}
#| echo: false
r3 <- min(cv_model3$results$RMSE)
```

RMSE `r round(r3, 3)`.

::: {style="color: lightblue"}
### Model 4: Ten Predictors
:::

```{r}
set.seed(123)
cv_model4 <- train(
  grade_sem2 ~ grade_sem1 + admission_grade + prev_grade + age + scholarship +
    fees_current + units_sem1 + Gender + Debtor + Displaced,
  data = train, method = "lm", trControl = ctrl
)
```

```{r}
#| echo: false
r4 <- min(cv_model4$results$RMSE)
```

RMSE `r round(r4, 3)`.

::: {style="color: lightblue"}
### Comparing Them on the Same Folds
:::

```{r}
summary(resamples(list(
  model1 = cv_model1,
  model2 = cv_model2,
  model3 = cv_model3,
  model4 = cv_model4
)))$statistics$RMSE
```

::: {.callout-important icon="false"}
## What the Table Shows

Going from one predictor to ten improved cross validated RMSE by roughly `r round(r1 - r4, 3)` grade points. On a 20 point scale that is a very small change.

Training fit never gets worse when you add a predictor. Predictive accuracy can, and here it barely changed even though several of those predictors are individually significant.

The [noise predictor example](week3_1.qmd) showed the same gap with random variables. Here it appears with real ones. This is the motivation for the regularization methods in the next chapter, which penalize predictors that do not improve prediction.
:::

::: {style="color: yellow"}
## Model Concerns
:::

Linear regression is widely used largely because its coefficients are easy to interpret. But it relies on several strong assumptions, and those assumptions are violated more often as predictors accumulate. Violations can produce biased estimates, misleading interpretations, and poor predictions.

::: {style="color: lightblue"}
### Linearity
:::

The model assumes the change in the outcome for a one unit change in a predictor is constant across the whole range of that predictor. Many real relationships are not like that.

Number of units passed in the first semester is a good example.

```{r}
p1 <- ggplot(train, aes(units_sem1, grade_sem2)) +
  geom_point(size = 1, alpha = .2) +
  geom_smooth(se = FALSE, color = "orange") +
  labs(x = "Units passed, first semester", y = "Second semester grade",
       title = "A curved relationship") +
  theme_minimal(base_size = 13)

p2 <- ggplot(train, aes(units_sem1, grade_sem2)) +
  geom_point(size = 1, alpha = .2) +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2), se = FALSE, color = "steelblue") +
  labs(x = "Units passed, first semester", y = "Second semester grade",
       title = "A squared term follows it") +
  theme_minimal(base_size = 13)

gridExtra::grid.arrange(p1, p2, nrow = 1)
```

A straight line through the left panel would underfit badly. Adding a squared term roughly doubles the variance this predictor accounts for:

```{r}
c(
  linear    = summary(lm(grade_sem2 ~ units_sem1, data = train))$r.squared,
  quadratic = summary(lm(grade_sem2 ~ units_sem1 + I(units_sem1^2), data = train))$r.squared
)
```

Transformations of the predictors, or of the outcome, are often enough to make a relationship approximately linear. The model is still a *linear model*, because it remains linear in its coefficients.

::: {style="color: lightblue"}
### Heteroskedasticity
:::

Heteroskedasticity means the variance of the errors is not constant across levels of the predictors, violating the homoscedasticity assumption.

```{r}
df1 <- broom::augment(cv_model1$finalModel, data = train)

ggplot(df1, aes(.fitted, .std.resid)) +
  geom_point(size = 1, alpha = .3) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "Predicted values", y = "Standardized residuals",
       title = "Model 1", subtitle = "grade_sem2 ~ grade_sem1") +
  theme_minimal(base_size = 13)
```

Look for a funnel shape, where the spread of residuals widens or narrows systematically as predictions get larger. Remedies include transforming the outcome or using robust standard errors.

Heteroskedasticity mainly affects **inference**, because the standard errors and p values covered in [Inference in Linear Regression](week3_1b_inference.qmd) assume constant variance. The predictions can still be accurate.

::: {style="color: lightblue"}
### Autocorrelation
:::

Autocorrelation occurs when residuals are correlated with each other, violating the independence assumption. It underestimates standard errors and produces overly optimistic p values. It matters most with a time component or with repeated measures on the same people. Each row here is a different student measured once, so we do not pursue it, though it becomes important with intensive longitudinal data.

::: {style="color: lightblue"}
### Feature Importance
:::

The `vip()` function (Variable Importance Plots) visualizes the relative importance of predictors. It accepts any fitted model object, including `lm()`, `glm()`, random forests, and boosted trees.

For linear models, `vip()` ranks predictors by the absolute value of their t statistics.

```{r}
vip(cv_model4, num_features = 10, method = "model")
```



Next: [Simple Logistic Regression](week3_3.qmd), which keeps this dataset and changes the outcome from a grade to a category.
