---
title: "Feature Engineering"
subtitle: "PSYC 559: Applied Machine Learning in Psychology"
author: "Your name here"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
    theme: flatly
---

```{r setup, include=FALSE}
# cache = TRUE matters here. The GSS download in the next section takes a
# minute or two, and caching means you only pay that cost on the first knit.
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      cache = TRUE)
```

This file is a standalone companion to the Feature Engineering unit on the
course site. It downloads its own data, so you can open it in RStudio and knit
it start to finish without any other files.

Feature engineering is everything you do to the predictors before a model ever
sees them. It is usually where most of the work in an applied project goes.

# Setup

```{r packages, cache=FALSE}
library(gssr)        # downloading the General Social Survey data
library(dplyr)       # data manipulation and the pipe
library(tidymodels)  # bundles recipes for preprocessing and rsample for splitting
library(caret)       # train() for fitting and cross validation
library(naniar)      # exploring and visualizing missing data
library(glmnet)      # the lasso model we fit at the end
```

# Building the Example Data

We use the [General Social Survey](https://gss.norc.org/), a long running,
nationally representative survey of adults in the United States that has run
almost every two years since 1972.

The outcome is `happy`, self reported happiness. The features are a set of
demographic and background variables we think might predict it.

**The next chunk downloads a large file, so give it a minute.**

```{r download}
gss24 <- gss_get_yr(2024)

target <- "happy"

features <- c(
  "age",     # age of participant
  "sex",     # sex of participant
  "race",    # race of participant
  "educ",    # highest education completed
  "income",  # income of participant
  "childs",  # number of children
  "wrkstat", # work force status
  "marital", # marital status
  "born",    # whether the participant was born in the USA
  "partyid", # political party
  "adults",  # number of family members 18 or older
  "earnrs"   # number of earners in the family
)

data <- gss24[, c(target, features)]

dim(data)
```

Now clean it up. Survey data arrives with its own missing value codes and with
labels that are not pleasant to read, so we fix both.

```{r clean}
cat_vars <- c("sex", "race", "educ", "wrkstat", "marital", "born", "partyid")
con_vars <- c("age", "income", "childs", "adults", "earnrs", "happy")

# Turns "MARRIED" into "Married". Splits each label on spaces, capitalizes the
# first letter of each word, lowercases the rest when strict = TRUE, then
# pastes the words back together.
capwords <- function(x, strict = FALSE) {
  cap <- function(x) paste(toupper(substring(x, 1, 1)),
                           {x <- substring(x, 2); if (strict) tolower(x) else x},
                           sep = "", collapse = " ")
  sapply(strsplit(x, split = " "), cap, USE.NAMES = !is.null(names(x)))
}

data <- data |>
  mutate(
    # The GSS codes missing values in several different ways. zap_missing turns
    # all of them into a plain NA so R treats them consistently.
    across(everything(), haven::zap_missing),
    across(all_of(cat_vars), forcats::as_factor),
    across(all_of(con_vars), as.numeric),
    across(all_of(cat_vars), \(x) forcats::fct_relabel(x, capwords, strict = TRUE))
  )

# Drop rows with no outcome. We cannot learn from a case whose answer is unknown.
data <- data[!is.na(data$happy), ]

dim(data)
```

# Part 1: Missing Data

Knowing how the missing data came about is critical for knowing how to handle
it. The usual framework distinguishes three mechanisms.

- **MCAR**, missing completely at random. Missingness is unrelated to anything,
  observed or unobserved. Rare in practice.
- **MAR**, missing at random. Missingness depends on variables you *did*
  observe. This is the assumption most modern methods rely on.
- **MNAR**, missing not at random. Missingness depends on the unobserved value
  itself, for example high earners declining to report income. This is the hard
  case and no imputation method fixes it on its own.

Start by looking at the missingness rather than guessing about it.

```{r vis-miss}
# Black marks a missing value, gray an observed one
vis_miss(data)
```

```{r miss-var}
# The same information as a ranked table
miss_var_summary(data)
```

## Mean Imputation

The simplest approach fills each gap with the column mean, or the mode for
categorical variables.

```{r impute-mean}
rec_mean <- recipe(happy ~ ., data = data) %>%
  step_impute_mean(all_numeric_predictors()) %>%
  step_impute_mode(all_factor_predictors())

# prep() learns the means from the data, bake() applies them.
# new_data = NULL is how you ask for the already prepped training data back.
data_mean <- prep(rec_mean) %>% bake(new_data = NULL)

vis_miss(data_mean)
```

It is fast and it always works, but it distorts the data. Every imputed value
sits exactly at the mean, which shrinks the variance of that variable and
weakens its relationship with everything else.

## KNN Imputation

A better approach uses the rest of the row. K nearest neighbor imputation finds
the cases most similar to the one with the gap and borrows from them.

Think of it this way. Somebody did not tell us whether they like pineapple.

- **Mean imputation** says most people like pineapple, so assume they do.
- **KNN imputation** says this person looks a lot like five other people who all
  like pineapple, so assume they probably do too.

The second uses information the first throws away.

```{r impute-knn}
rec_knn <- recipe(happy ~ ., data = data) %>%
  step_impute_knn(all_predictors())

data_knn <- prep(rec_knn) %>% bake(new_data = NULL)

vis_miss(data_knn)
```

> **Your turn.** Compare the standard deviation of `income` before imputation,
> after mean imputation, and after KNN imputation. Which one changes it least,
> and why would you expect that?

```{r your-turn-1}
# Your code here

```

# Part 2: Feature Filtering

Filtering evaluates features *before* fitting and drops the ones that cannot
help. The goals are fewer dimensions, less redundancy, easier interpretation,
and better generalization.

The most common filter removes features with almost no variation. A variable
that is the same value for nearly everybody carries almost no information, and
it can break a model outright if a rare level lands entirely in one fold.

```{r nzv}
# saveMetrics = TRUE returns the diagnostics for every variable rather than
# just the column numbers of the ones flagged for removal
caret::nearZeroVar(data, saveMetrics = TRUE) %>%
  tibble::rownames_to_column()
```

Two columns matter.

- `freqRatio` is the frequency of the most common value divided by that of the
  second most common. A large value means one level dominates.
- `percentUnique` is the number of distinct values as a percentage of rows. A
  small value means the variable is nearly constant.

Inside a recipe this becomes one step.

```{r step-nzv}
recipe(happy ~ ., data = data) %>%
  step_nzv(all_predictors())
```

# Part 3: Numeric Features

## Skewness

Models with distributional assumptions can benefit from reducing skew. The
Yeo-Johnson transformation is a flexible option that, unlike Box-Cox, handles
zero and negative values.

```{r yeojohnson}
# all_numeric_predictors() keeps the transformation off the outcome. A recipe
# only records the plan here, nothing is computed until it is prepped.
recipe(happy ~ ., data = data) %>%
  step_YeoJohnson(all_numeric_predictors())
```

## Standardization

Standardizing puts features on a common scale, mean zero and standard deviation
one. This keeps variables with large numeric ranges, like income in dollars,
from dominating variables with small ones, like number of children. It matters
for any distance based method and for penalized regression, where the penalty
would otherwise fall unevenly across predictors.

```{r standardize}
# step_center subtracts the mean, step_scale divides by the standard deviation.
# Both learn those numbers from the training data alone, which is what keeps
# information from the test set out of the calculation.
recipe(happy ~ ., data = data) %>%
  step_center(all_numeric_predictors()) %>%
  step_scale(all_numeric_predictors())
```

# Part 4: Categorical Features

## Lumping

Some features have levels with very few observations. Look at work status.

```{r wrkstat}
count(data, wrkstat) %>% arrange(n)
```

Several levels are rare enough that a model cannot estimate anything stable for
them. `step_other` collapses everything below a threshold into a single
category.

```{r lumping}
lumping <- recipe(happy ~ ., data = data) %>%
  step_other(wrkstat, threshold = 0.05, other = "Other")

data_lumped <- prep(lumping, training = data) %>% bake(new_data = NULL)

count(data_lumped, wrkstat) %>% arrange(n)
```

## Dummy Encoding

Most models require numeric predictors, so categories have to become numbers.
Dummy encoding turns a factor with k levels into indicator columns.

```{r dummy}
# one_hot = TRUE makes one column per level. FALSE drops one to serve as the
# reference category, which is what regression models normally want.
recipe(happy ~ ., data = data) %>%
  step_dummy(all_factor_predictors(), one_hot = TRUE)
```

Tree based models are an exception and can handle factors directly, though they
often still benefit from lumping.

# Part 5: The Whole Workflow

Now put it together. Split first, write one recipe, fit with cross validation.

```{r workflow-split}
set.seed(123)
split <- initial_split(data, prop = 0.7, strata = "happy")

data_train <- training(split)
data_test  <- testing(split)
```

The recipe collects every step from the sections above into one blueprint.
Order matters. Filtering comes before imputation so we do not spend effort
imputing a feature we are about to drop, and dummy encoding comes before
standardizing so the resulting indicator columns are on the same scale too.

```{r workflow-recipe}
blueprint <- recipe(happy ~ ., data = data_train) %>%
  step_nzv(all_nominal()) %>%
  step_impute_bag(all_predictors()) %>%
  step_dummy(all_factor_predictors(), one_hot = FALSE) %>%
  step_center(all_numeric(), -all_outcomes()) %>%
  step_scale(all_numeric(), -all_outcomes())

blueprint
```

Now fit. The model is the **lasso**, a linear regression that pays a penalty for
large coefficients. `lambda` sets how steep that penalty is, and it is a
hyperparameter, so cross validation picks it rather than us.

```{r workflow-fit}
cv <- trainControl(method = "cv", number = 10)

# Candidate penalties, spaced evenly on a log scale rather than a linear one,
# because what matters is the order of magnitude of the penalty.
hyper_grid <- expand.grid(
  alpha  = 1,  # alpha = 1 is lasso, alpha = 0 would be ridge
  lambda = 10^seq(-4, -1.5, length.out = 10)
)

set.seed(123)
lasso_fit <- train(
  blueprint,
  data = data_train,
  method = "glmnet",
  trControl = cv,
  tuneGrid = hyper_grid,
  metric = "RMSE"
)

lasso_fit
```

```{r workflow-plot}
ggplot(lasso_fit)
```

Cross validated error falls at first, because a little shrinkage removes noise
the model was chasing, then turns and climbs once the penalty is strong enough
to flatten coefficients carrying real signal. The bottom of that curve is the
value `caret` keeps.

Notice that the recipe went into `train()` rather than being applied to
`data_train` beforehand. That is deliberate. It means the recipe is re-learned
inside every fold, using only that fold's training portion, so no information
crosses from a held out fold into the fitting.

# Exercises

**1.** Swap `step_impute_bag()` for `step_impute_mean()` in the blueprint and
refit. Does cross validated RMSE change much? What does that tell you about how
much the imputation choice is doing here?

```{r ex-1}
# Your code here

```

**2.** Add `step_YeoJohnson(all_numeric_predictors())` to the blueprint before
the centering steps and refit. Did reducing skew help?

```{r ex-2}
# Your code here

```

**3.** Add `step_other(all_nominal_predictors(), threshold = 0.05)` to lump rare
levels across every categorical feature, not just `wrkstat`. Compare the number
of columns the recipe produces before and after.

```{r ex-3}
# Your code here

```

**4.** Take the fitted model to the test set, once, using `postResample()`. How
does the test RMSE compare to the cross validated RMSE, and which of the two
would you report?

```{r ex-4}
# Your code here

```
