---
title: "The Modeling Pipeline"
subtitle: "PSYC 559: Applied Machine Learning in Psychology"
author: "Your name here"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

This file is a standalone companion to the Modeling Pipeline unit on the course
site. Everything it needs is downloaded or built inside the file, so you can
open it in RStudio and knit it start to finish without any other files.

Work through it in order. Run each chunk with the green arrow at its top right,
or knit the whole document to see all of the output at once.

# Setup

Load these six packages. All of them came with the course setup, so you should
not need to install anything. If a package is missing, install it once with
`install.packages("name")` and then run this chunk again.

```{r packages}
library(rsample)      # splitting data into training and test sets
library(recipes)      # writing preprocessing steps down as a recipe
library(caret)        # fitting models and running cross validation
library(AmesHousing)  # the example data
library(dplyr)        # data manipulation and the pipe
library(glmnet)       # the lasso model we fit near the end
```

# The Data

We use the **Ames housing** data throughout this unit.

- Every sale of a residential property in Ames, Iowa between 2006 and 2010.
- 2,930 houses and 81 variables, so each row is one sale.
- The target is `Sale_Price`, in dollars.
- The predictors describe the house rather than the buyer, including size, age,
  quality ratings, and which of 28 neighborhoods it sits in.
- There are no missing values, which is unusual and makes it clean for teaching.

```{r data}
ames <- make_ames()

dim(ames)
summary(ames$Sale_Price)
```

The mean sits above the median because a handful of expensive houses pull the
distribution to the right. That skew matters twice below, once for stratified
splitting and once for the gap between RMSE and MAE.

# Part 1: Splitting

The `rsample` package handles splitting. Three functions do nearly all of it.

- `initial_split()` decides which rows go where.
- `training()` pulls out the training rows.
- `testing()` pulls out the test rows.

```{r split-plain}
# set.seed() comes first because the split is random. Without a seed you get a
# different split on every run and none of your numbers reproduce.
set.seed(559)

split <- initial_split(ames, prop = 0.7)

ames_train <- training(split)
ames_test  <- testing(split)

c(train = nrow(ames_train), test = nrow(ames_test))
```

There is nothing magic about `prop = 0.7`. Common choices run from 0.6 to 0.8.
More training data gives a better model, more test data gives a steadier
estimate of how good it is.

## Stratified Splitting

A plain random split can get unlucky. If the expensive houses land mostly in one
half, the two halves are not comparable. `strata` splits *within* groups of the
outcome and then recombines, so both halves end up looking alike.

```{r split-strata}
set.seed(559)

split <- initial_split(ames, prop = 0.7, strata = "Sale_Price")

ames_train <- training(split)
ames_test  <- testing(split)

c(train_mean = mean(ames_train$Sale_Price),
  test_mean  = mean(ames_test$Sale_Price))
```

The two means now sit close together by construction rather than by luck.

> **Your turn.** Change the seed in the unstratified chunk a few times and watch
> the two means move around. Then do the same with `strata`. How much does
> stratifying reduce the wobble?

```{r your-turn-1}
# Your code here

```

## Cross Validation

`trainControl()` describes *how* to resample. `train()` does the fitting.

```{r cv-setup}
cv <- trainControl(
  method = "cv",  # k fold cross validation
  number = 10     # 10 folds
)
```

That says: split the training data into 10 folds, hold each fold out in turn,
fit on the other 9, and score on the fold that was held out.

```{r cv-fit}
set.seed(559)

fit_cv <- train(
  Sale_Price ~ Gr_Liv_Area,
  data      = ames_train,
  method    = "lm",
  trControl = cv
)

fit_cv
```

The reported RMSE is the **average across the 10 folds**, not one number from
one split. A single holdout gives an unstable estimate. Averaging ten of them
gives a steadier one.

`train()` hides a lot, so it is worth being explicit about what just happened.

1. The training data were divided into 10 folds.
2. The model was fit 10 separate times, each time on 9 folds.
3. Each fit was scored on the fold it had not seen.
4. Those 10 scores were averaged.
5. Finally a model was fit once on all of the training data. That is the model
   you keep.

Note what did **not** happen. The test set was never touched. It is still
sitting untouched in `ames_test`, which is where it stays until the very end.

# Part 2: Evaluating Performance

Every regression metric starts in the same place, the difference between what
actually happened and what the model said would happen.

$$e_i = y_i - \hat{y}_i$$

```{r residuals}
fit <- lm(Sale_Price ~ Gr_Liv_Area, data = ames_train)

predicted <- predict(fit, ames_test)
actual    <- ames_test$Sale_Price
residuals <- actual - predicted

head(round(residuals))
```

A positive residual means the house sold for more than we predicted. A negative
one means we predicted too high.

## RMSE and MAE

$$RMSE = \sqrt{\frac{1}{n}\sum(y_i - \hat{y}_i)^2}
\qquad
MAE = \frac{1}{n}\sum|y_i - \hat{y}_i|$$

```{r rmse-mae}
rmse <- sqrt(mean(residuals^2))  # square, average, then undo the squaring
mae  <- mean(abs(residuals))     # drop the sign, then average

c(RMSE = rmse, MAE = mae)
```

Both read in dollars, but they punish errors differently.

- **RMSE** squares first, so one error of 100,000 counts a hundred times as much
  as an error of 10,000, not ten times. Large misses dominate it.
- **MAE** treats errors proportionally. An error twice as large counts twice.

RMSE is the larger of the two and always will be unless every residual is
identical. The gap between them is driven by a few houses the model got badly
wrong.

```{r worst-misses}
sort(abs(residuals), decreasing = TRUE)[1:5]
```

Those five are pulling RMSE up. MAE barely notices them.

Which one you report is a real decision. If a few catastrophic misses matter far
more than many small ones, RMSE is the honest summary. If you are describing
typical accuracy to somebody, MAE answers the question they actually asked.
Neither is more correct. They answer different questions.

## R Squared

$$R^2 = 1 - \frac{\sum(y_i - \hat{y}_i)^2}{\sum(y_i - \bar{y})^2}$$

The bottom is how wrong you would be if you ignored the predictors entirely and
guessed the mean every time. The top is how wrong your model is.

```{r r2}
r2 <- 1 - sum(residuals^2) / sum((actual - mean(actual))^2)
r2
```

It is popular because it is unitless, which makes it easy to compare across
problems. It is also easy to over read. Our model explains about half the
variation in price and is still off by tens of thousands of dollars.

`caret` computes all three at once, so you rarely do this by hand.

```{r postresample}
postResample(pred = predicted, obs = actual)
```

> **Your turn.** Your model has an R squared of 0.51. A colleague reports 0.78 on
> a different dataset with a different outcome. Whose model is better? Write two
> sentences below.

*Your answer here.*

# Part 3: The Whole Pipeline

Now all six steps in order, on one worked example.

## Step 1: Split

```{r pipe-split}
set.seed(559)
split <- initial_split(ames, prop = 0.7, strata = "Sale_Price")

ames_train <- training(split)
ames_test  <- testing(split)
```

From here until step 6, `ames_test` does not appear anywhere.

## Step 2: A Preprocessing Recipe

A recipe is a written list of preprocessing steps that is *learned from the
training data and then applied elsewhere*.

```{r pipe-recipe}
rec <- recipe(Sale_Price ~ Gr_Liv_Area + Year_Built + Neighborhood,
              data = ames_train) %>%
  step_other(all_nominal_predictors(), threshold = 0.05) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())

rec
```

Reading it in order:

- `step_other` lumps rare neighborhoods into a single "other" category, since a
  neighborhood appearing four times cannot support its own estimate.
- `step_normalize` centers and scales the numeric predictors.
- `step_dummy` converts categories into 0 and 1 indicator columns.
- `step_zv` drops any predictor with zero variance.

**Why a recipe rather than just doing it.** Suppose you center and scale
`Gr_Liv_Area` before splitting. The mean you subtract is computed from *all* the
data, including rows destined for the test set. Information from the test set
has now leaked into training and your test estimate is optimistic. A recipe
prevents this by construction. It records that a mean must be subtracted, learns
that mean from training data only, and applies that stored value to new data
later. Nothing errors when you get this wrong. Your numbers just quietly become
too good.

## Step 3: Resampling Scheme

```{r pipe-cv}
cv <- trainControl(method = "cv", number = 10)
```

## Step 4: Fit

Passing the recipe to `train()` rather than a formula means the recipe is
re-learned inside every fold, on that fold's training portion only. That is what
keeps the leakage protection honest during cross validation.

```{r pipe-lm}
set.seed(559)

fit_lm <- train(
  rec,
  data      = ames_train,
  method    = "lm",
  trControl = cv,
  metric    = "RMSE"
)

fit_lm$results
```

## Step 5: Compare Candidates

We cannot know in advance which method wins, so we try more than one and compare
them on the same folds.

Here is a **lasso** model, a linear regression that pays a penalty for large
coefficients. `lambda` sets how steep that penalty is, and it is a
hyperparameter, so we do not pick it ourselves. We hand `caret` a grid of
candidates and cross validation chooses.

```{r pipe-lasso}
set.seed(559)

fit_lasso <- train(
  rec,
  data      = ames_train,
  method    = "glmnet",
  trControl = cv,
  metric    = "RMSE",
  # alpha = 1 is what makes this lasso rather than one of the other
  # penalized models glmnet can fit
  tuneGrid  = expand.grid(alpha = 1, lambda = c(1, 10, 100, 1000, 10000))
)

fit_lasso$results[, c("lambda", "RMSE", "Rsquared", "MAE")]
```

Look at the shape of that RMSE column. Small penalties barely move the model, so
those rows are nearly identical. Once the penalty gets large enough it starts
squashing coefficients that were doing useful work, and the error climbs.

```{r pipe-best}
fit_lasso$bestTune
```

Because both models used the same seed and the same folds, comparing them is
fair.

```{r pipe-compare}
results <- resamples(list(linear = fit_lm, lasso = fit_lasso))
summary(results)$statistics$RMSE
```

The two are essentially tied, and that is a real result rather than a broken
comparison. We gave the model three predictors and roughly two thousand training
rows, so there is nothing for a penalty to clean up. Penalties start to matter
when there are many predictors relative to rows.

**Notice what chose the winner.** The validation folds picked the value of
`lambda`, and the validation folds picked between the two models. Both are
decisions made *with data*, which is exactly why they cannot be made using the
test set.

## Step 6: The Test Set, Once

```{r pipe-test}
final_model <- fit_lm  # the two tied, so keep the simpler one

test_predictions <- predict(final_model, ames_test)

postResample(pred = test_predictions, obs = ames_test$Sale_Price)
```

That is the number you report. It estimates how the model will perform on houses
nobody has seen.

If you now go back, change the model, and re-run this, the number stops meaning
what it meant. The test set has become part of model selection and your estimate
is optimistic again. You get one look. This is why steps 1 through 5 exist.

# Exercises

Use the chunks below. The pipeline shape stays the same, you are only changing
one piece at a time.

**1.** Refit the linear model with more predictors. Add `Total_Bsmt_SF` and
`Garage_Cars` to the recipe. Does cross validated RMSE improve?

```{r ex-1}
# Your code here

```

**2.** Now that the model has more predictors, rerun the lasso comparison. Does
the penalty do more work than it did with only three predictors?

```{r ex-2}
# Your code here

```

**3.** Change `metric = "RMSE"` to `metric = "MAE"` in one of the fits. Does
`train()` select a different model? Explain why it might.

```{r ex-3}
# Your code here

```

**4.** Deliberately break the pipeline. Normalize `Gr_Liv_Area` across the full
`ames` dataset *before* splitting, then fit and evaluate. Is the test RMSE
better or worse than the honest version, and why is "better" the wrong thing to
be pleased about here?

```{r ex-4}
# Your code here

```
